---
title: "BUAN 6530 Business forecasting and predictive analytic"
author: "Mark Drummond"
date: "Fall 2020 - Final Projet"
output:
  pdf_document:
  word_document:
  html_document:
     df_print: paged
---

```{r}
library(fpp2)
```

```{r}
# Display Working Directory and System Time
getwd()
Sys.time()
```

```{r}
#You may run the code by deleting "#" below the 1st time if you cannot knit a pdf file
#tinytex::install_tinytex()
```

```{r}
#install.packages("tseries")
#install.packages("quantmod")
#install.packages("ggplot2")
#install.packages("xlsx")
#library(tseries)
#library(quantmod)
#library(ggplot2)
#library(forecast)
library(fma)
library(Mcomp)
library(smooth)
 
```
##  Load data from File
### 
### 

```{r}
# a
# Monthly total Deaths from 1/1/1999 to 10/1/2020
deathsx <- readxl::read_excel("Data Sets/ALL Deaths 1999 2020.xlsx") 

```

```{r}
# b
head(deathsx)

plot(deathsx$MonthYear, deathsx$Deaths, col="blue", type = "l", lwd=1)

```

```{r}
# Convert the data into time series
# limit to before any Covid Deaths were reported
death_ts <- ts(deathsx[,-1], start=1999,end=c(2020,2),  frequency=12)
#death_ts <- ts(deathsx[,-1], frequency=12)
#  Name death_ts[, ”Deaths”] as ‘death’
deaths <- death_ts[, 'Deaths']

```

```{r}
#head(deaths)
#View(deaths)
```

```{r}
# Plot ‘death’ and identify  any seasonal pattern and stationarity
autoplot(deaths, xlab = 'Years', ylab = "Deaths")

```

```{r}
# look at the seasonal plot
ggseasonplot(deaths)

```

```{r}
```


```{r}
# stationarity,ACF and PACF
ggtsdisplay(deaths)

# Add additional commentary here
# The PACF plots show that there are significant lags at 1 and 13.
# The AFC plot there is a geometric decay at each Lag. 
# This would indicate that a seasonal AR model would be used

```


```{r}
# Get Box Cov
BoxCox.lambda(deaths)
```
  
```{r}
# For Non Seasonal
ndiffs(BoxCox(deaths,lambda=0.250792))
# = 1 so the model needs a non-Seasonal differencing of 1

```


```{r}
# For Seasonal
# Gives me d for the Arima Model
# Difference once based on ndiff 
nsdiffs(diff(BoxCox(deaths,lambda=0.250792),lag=12))
# = 0 so no seasonality needed

```

```{r}
ggtsdisplay(diff(deaths,lag=12))
# Looks stationary now
# The AFC plot there is a geometric decay 

# The PACF plots show that there are significant lags at 1 then minor ones at 12,13,24,25.
# This would indicate that a AR model with some seasonality should be used

```

```{r}
#   Comments on the seasonality, stationarity, and possible forecasting strategies
#   Appears to have yearly seasonality and a trend so you should take 1 difference 
#   with the 1st difference then
#   Looks to be following a ARIMA(1,d,q) since
#   The spikes on in the ACF and PACF drops significantly after lag 12
#   and most of the other spikes are with the blue lines in the ACF plot
```


# Section 2. Testing diffrent models 
# 
## Section 2.1 
###  Simple Moving Averages (SMA)

```{r}
# a
dea_sma1 <- sma(deaths,lambda=0.250792)
dea_sma2 <- sma(deaths,order=12,lambda=0.250792)

```

```{r}
# 
# Summarize the sma models results and explain the parameters of the optimal model
accuracy(dea_sma1$fitted,deaths)
accuracy(dea_sma2$fitted,deaths)


summary(dea_sma1)
summary(dea_sma2)

# MODEL1 HAVE THE lowest RSME 13252.08   vs 14255.52 
checkresiduals(dea_sma1)
# STILL SOME SPIKES outside the blue lines

```


```{r}
###  Exponential Smoothing  (SES)
#
dea_ses1 <- ses(deaths, lambda=0.250792)
dea_ses2 <- ses(deaths, lambda=0.250792,alpha=0.1)

```

```{r}
# 
accuracy(dea_ses1$fitted,deaths)
#accuracy(dea_ses2$fitted,deaths)

summary(dea_ses1)
#print('####################')
#summary(dea_ses2)
  # BEST MODEL IS dea_ses1 with a RMSE 13246.03  vs 14611.42    
checkresiduals(dea_ses1)

```
  
```{r}
# g
# SES 1 is the best model with a RMSE = 13978.23
#dea$ses_fit <- dea_ses1$fitted
#head(dea)

```

## Section 2.2 - Simple Linear Regression
#
### Time Series with TREND
## 


```{r}
#print('#################### With TREND #######################')
# a
deaths_tslm1 <- tslm(data=death_ts,deaths ~ trend )

```

```{r}
# 
summary(deaths_tslm1)
accuracy(deaths_tslm1)
# RMSE 14866.79  Trend t-value is 11.72 is above 2 and so a good for the model
# Adjusted R-squared: 0.3052 
# R-squared:  0.3456

```

### Time Series with TREND & SEASON
## 

```{r}
#print('#################### With TREND & SEASON #######################')
##
deaths_tslm2 <- tslm(data=death_ts,Deaths ~ trend + season )

```


```{r}
# 
summary(deaths_tslm2)
accuracy(deaths_tslm2)
# Check the model
checkresiduals(deaths_tslm2)
# RMSE 9222.548   All of the seasons have t-values of above 2 which is good
# Adjusted R-squared:  0.758 
# R-squared:  0.8271

```


```{r}
# 
# With TREND and Season has the lowest Adjusted R-squared: 0.758, RMSE = 9222.548 
# And of the seasons have t-values of above 2 
# 
#dea$tslm2_fit <-deaths_tslm2$fitted.values

#head(dea) # to see if it works
```

## Section 2.3 - Holt, HW and ETS
#
### Holt's method 
##

```{r}
# a - HOLT
#print('#################### HOLT #######################')
deaths_holt <- holt(deaths,lambda=0.250792)
```

```{r}
# b
summary(deaths_holt)
accuracy(deaths_holt)

# RMSE 13247.62 

```


#
### Holt-Winters' additive method 

```{r}
# c  HW
#print('#################### HW #######################')
deaths_hw <- hw(deaths,seasonal="additive")
```

```{r}
# d
summary(deaths_hw)
accuracy(deaths_hw)

# RMSE  5186.121 

```

#
### Holt-Winters' multiplicative method

```{r}
# multiplicative
deaths_hw <- hw(deaths,seasonal="multiplicative")
```

```{r}
# d
summary(deaths_hw)
accuracy(deaths_hw)

# RMSE  5535.074  

```



### Exponential smoothing (ETS)
#
```{r}
# e ets
#print('#################### ETS #######################')
deaths_ets <- ets(deaths)

```

```{r}
# f
summary(deaths_ets)
accuracy(deaths_ets)

# RMSE 5037.35 
```

```{r}
# g
# ETS had the lowest RSM value of 5037.35   
#dea$ets_fit <- dea_ets$fitted
#head(dea) # to see if it worked
```



## Section 2.4 - ARIMA MODELS
###
##

```{r}
BoxCox.lambda(deaths)
```

```{r}
# For Non Seasonal
ndiffs(BoxCox(deaths,lambda=0.250792))
# Gives me d for the Arima Model
# = 1 so the model needs a non-Seasonal differencing of 1

```
```{r}
# For Seasonal
# Gives me d for the Arima Model
nsdiffs(diff(BoxCox(deaths,lambda=0.250792),lag=12))
# = 0 so do not need to Seasonal differencing  of 1 after one Diff indicated in ndiff
```

```{r}
# 
#	The PACF plots show that there are significant lags at 1 then minor ones at 12,13,24,25. 
#	The AFC plot there is a geometric decay at lag 12. This would indicate that a seasonal AR model should also be tried.

ggtsdisplay(diff(BoxCox(deaths,lambda=0.250792),lag = 12))

```


### ARIMA Manual
#
```{r}
# Manual ARIMA

#dea_arima1 <- Arima(deaths, order = c(2,0,0),seasonal=c(2,1,1),lambda=0.250792)
# AICc = 386.65  

#dea_arima1 <- Arima(deaths, order = c(1,0,1),seasonal=c(2,1,1),lambda=0.250792)
# AICc = 385.7

#dea_arima1 <- Arima(deaths, order = c(2,0,2),seasonal=c(2,1,0),lambda=0.250792)
# AICc = 382.31

#dea_arima1 <- Arima(deaths, order = c(2,0,1),seasonal=c(1,1,1),lambda=0.250792)
# AICc = 373.47

#dea_arima1 <- Arima(deaths, order = c(2,0,2),seasonal=c(2,1,1),lambda=0.250792)
# AICc = 369.15

#dea_arima1 <- Arima(deaths, order = c(1,0,2),seasonal=c(2,1,2),lambda=0.250792)
# AICc = 369.14   

dea_arima1 <- Arima(deaths, order = c(2,0,2),seasonal=c(1,1,1),lambda=0.250792)
# Best Model LOWEST AICc
# AICC = 368.60
# RMSE = 4787.379

summary(dea_arima1)
checkresiduals(dea_arima1)

```   

### AUTO ARIMA
#

```{r}
# AUTO ARIMA
aarima <- auto.arima(deaths , lambda=0.250792)
summary(aarima)

# AICc = 369.15   
# RMSE = 4914.212   
# Not better that the manual 
```

```{r}
# f The auto Arima Model had a RMSE value of 1.587109    
#   The Arima Model of order = c(0,1,0),seasonal=c(1,1,0) has a RMSE = 49.51 
#   They match

# Check the 
checkresiduals(aarima)

```
## Section 3 - Forecast Best Model
#
### ARIMA
##
```{r}
# g
# Manual Arima model has the lowest RSM value of 44.3173
#deaths$arima1_fit <- dea_arima1$fitted

#head(deaths) # to see if it worked
#head(dea_arima1$fitted)
autoplot(forecast(dea_arima1))
forecast(dea_arima1)
forecast(dea_arima1$fitted)


```
```{r}
#deathForAct <- readxl::read_excel("Data Sets/ALL Deaths 1999 2020.xlsx") 
deathForAct <- readxl::read_excel("Data Sets/FinalProjet Forcast Vs Actual-Mark Drummond.xlsx")
```

```{r}
# DEATHS with ACTUALS
death_tsAA <- ts(deathsx[,-1], frequency=12)
deathsAAA <- death_tsAA[, 'Deaths']
#View(deathsAAA)

```

```{r}
ggplot(data = deathForAct, aes(x=Month, y=Hi.95)) + geom_line(aes(colour=Actual))
```

```{r}
#Plot Actuals
plot(deathForAct$Month, deathForAct$Actual ,col="blue", type="l"  ,xlab="Date", ylab="Deaths", main="Actual vs Forecasted Deaths", ylim=c(200000,310000))
points(deathForAct$Month, deathForAct$Hi.95, col="red", type="l")
legend("topright", c("Actual", "Forecast"), col = c("blue", "red"),
       text.col = "black", lty = c(1, 1), pch = c("-","-"),
       merge = TRUE, bg = "gray90")


```



```{r}
# Write out forcast into ‘examreview-your-name.csv’ file
write.csv(forecast(dea_arima1$fitted),"Data Sets/FinalProjetForecast-Mark Drummond.csv")

```




